Rao-Blackwellised Two-filter algorithm for Conditionally Linear and Gaussian models

### 2016-11-07 TexArxiv/clgm.tex
Changes in Forward filter

\begin{abstract}
This paper focuses on Sequential Monte Carlo based approximations of smoothing distributions in conditionally linear and Gaussian state spaces. Only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps for these models. The most widely used Rao Blackwellised Sequential Monte Carlo smoothers introduced in \cite{kim:1994,lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} are based on the Forward Backward decomposition of the marginal smoothing distribution. In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps without any approximation in both forward and backward passes. In this paper, these algorithms are compared to an extension of the Two-filter based algorithm of \cite{briers:doucet:maskell:2010} with additional explicit Rao Blackwellised steps to improve its performance. The procedures are applied to commodity markets which are described using a two factor model based on the spot price and a convenience yield. Parameters are estimated using  an Expectation Maximization and the performance of the proposed procedure is assessed with Chicago Mercantile Exchange crude oil data.

@@

An important feature of these models is that, conditionally on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian. To exploit this specific structure, it has been suggested in \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to combine Sequential Monte Carlo (SMC) to sample indicator variables with the Kalman filter leading to Rao Blackwellised Monte Carlo filters. %These SMC methods based on sequential importance sampling and resampling techniques introduced in \cite{handschin:mayne:1969} and \cite{rubin:1987} allow to approximate any sequence of conditional distributions of the hidden states given the observations, see \cite{kitagawa:1987,gordon:salmond:smith:1993}. 
Improvements of these early filtering techniques  have been introduced in \cite{doucet:gordon:krishnamurthy:2001,schon:gustafsson:nordlund:2005}. %These algorithms sample only trajectories of indicator variables associated with importance weights, the corresponding state variables $(Z_1,\ldots,Z_k)$ are marginalized and estimated with a Kalman filter, which is the optimal filter in this case. 
%However, many maximum likelihood based estimation procedures such as  the Expectation Maximization  (EM)  algorithm introduced in \cite{dempster:laird:rubin:1977} require to approximate some marginal smoothing distributions of the states and indicator variables. 
%Several algorithms have been developed to solve the smoothing problem using SMC methods: for instance the Forward Filtering Backward Smoothing algorithm (FFBS), the Forward Filtering Backward Simulation algorithm (FFBSi) and the PArticle-based, Rapid Incremental Smoother (PaRIS)  developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000},  \cite{godsill:doucet:west:2004} and \cite{olsson:westerborn:2015}. 
%,lindsten:schon:2011
%In the forward pass of these algorithms, a Rao Blackwellised particle filter such as \cite{chen:liu:2000} is used to approximate the filtering distributions and the corresponding particles, weights and associated trajectories of means and variances of the Kalman filters are stored.
The use of Rao Blackwellisation to solve the smoothing problem has been proved to be challenging and has received satisfactory solutions only recently. The first forward-backward smoother proposed in the literature \cite{fong:godsill:doucet:west:2002} was not fully Rao Blackwellised as it required to sample the hidden linear states in the backward pass. An alternative, based on a structural approximation of the model suggested in \cite{kim:1994}, was introduced by \cite{barber:2006} to handle the backward pass.  The first fully Rao Blackwellised SMC smoother which does use this strong approximation of the backward pass was proposed by  \cite{briers:doucet:maskell:2010} and extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.  This two-filter approach combines a forward and a backward filters which are approximated numerically using SMC and Kalman filtering techniques. More recently, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} introduced a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions. 
%The backward pass samples trajectories of indicator variables backward in time using the particles and weights of the forward pass.

In this paper, we propose to improve the performance of the algorithms introduced in \cite{briers:doucet:maskell:2010} and in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} by using additional Rao Blackwellisation steps to produce more robust approximations.  In \cite{briers:doucet:maskell:2010}, for all $1\le i\le n$, the sampled forward and backward sequences are merged to approximate the posterior distribution of $(a_i,z_i)$. We propose to approximate the marginal smoothing distribution of $(a_i,z_i)$ by merging the sampled forward and backward trajectories at times $i-1$ and $i+1$ and integrating out all possible paths between time $i-1$ and time $i$ and between time $i$ and time $i+1$ instead of sampling random variables. Similarly, in the backward pass of \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013}, an indicator variable $\tilde{a}_i$ is sampled at time $1\le i\le n-1$ using the particles produced by the forward filter at time $i$. In this case, particle rejuvenation may be introduced by using the forward weighted samples at time $i-1$ and extending these trajectories at time $i$ with a Kalman filter for all possible values of the indicator variable. Then, $\tilde{a}_i$ is sampled in $\{1,\ldots,J\}$ using an appropriately adapted weight.

The paper is organized as follows. An overview of the proposed Monte Carlo algorithms to estimate the smoothing distributions of the hidden states as long as the additional explicit marginalization steps are presented in Section~\ref{sec:RaoBlackwell}. The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data. Then, a new model is proposed to explain regime switchings for commodity markets. The parameters of the proposed model are estimated using a Monte Carlo EM algorithm where the required conditional expectations are replaced by their estimates based on the proposed SMC method.  
The performance of the extended Rao-Blackwellised two-filter  algorithm is assessed using simulated data and CME crude oil (WTI) data. Intermediate computations and explicit approximations provided by the algorithm are given in the Appendix.

\section{Rao Blackwellised smoothing algorithms}
\label{sec:RaoBlackwell}
This section details the Sequential Monte Carlo algorithms which can be used to approximate the conditional distribution of the states $(a_1,\ldots,a_n)$ or the marginal distributions of $(a_i,z_{i})$ given the observations $(Y_1,\ldots,Y_n)$. For all $m\,\times\,m$ positive-definite matrix $A$ let $|A|$ be the determinant of $A$ and  for all $z\in\mathbb{R}^{m}$ define
\[
\normMat{A}{z}^2 \eqdef z'A^{-1}z\eqsp.
\]
Let $m_{}(a_{i},z_{i-1};z_i)$ be the probability density of the conditional distribution of $Z_i$ given $(a_{i},Z_{i-1})$ and $g_{}(a_{i},z_{i};y_i)$ be the probability density of the conditional distribution of $Y_i$ given $(a_{i},Z_{i})$:
\begin{align}
\label{eq:definition-m}
m_{}(a_{i},z_{i-1};z_i) & \eqdef \left|2\pi\barH_{a_i}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\barH_{a_i}}{z_i -d_{a_{i}} -T_{a_{i}}z_{i-1}}^2\right\}\eqsp,\\
\label{eq:definition-g}
g_{}(a_{i},z_{i};y_i) & \eqdef \left|2\pi\barG_{a_i}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\barG_{a_i}}{y_i - c_{a_i} - B_{a_i}z_{i}}^2\right\}\eqsp,
\end{align}
where
\[
\barG_j \eqdef G_jG'_j\;,\; \barH_{j} \eqdef H_jH'_j\eqsp.
\]
%Maximum likelihood estimation of the parameters using the Expectation Maximization algorithm or gradient ascent algorithms requires to compute the posterior distributions of the states $a_{i-1:i},z_{i-1:i}$  given the observations $Y_{1:n}$, written $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$, which cannot be computed in closed-form (the dependency of $\phi^n_{i}$ on $Y_{1:n}$ is dropped for better clarity). SMC methods provide a class of importance sampling and resampling procedures to approximate $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$. 
All the algorithms considered in this paper are based on forward-backward or two-filter decompositions of the smoothing distributions and share the same forward filter detailed in Section~\ref{sec:forward}.

\subsection{Forward filter}
\label{sec:forward}
The SMC approximation $p^N_{}(a_{1:i},z_{i}|y_{1:i})$ of $p_{}(a_{1:i},z_{i}|y_{1:i})$ may be obtained using the standard Rao Blackwellised  algorithm given in \cite{chen:liu:2000}. The procedure produces a sequence of trajectories $(a^k_{1:i})_{1\le k \le N}$ associated with importance weights $(\omega^k_{i})_{1\le k \le N}$ used to define the following approximation of $p_{}(a_{1:i},z_{i}|y_{1:i})$:
\[
p^N_{}(a_{1:i},z_{i}|y_{1:i}) = \sum_{k=1}^N\delta_{a^k_{1:i}}(a_{1:i})\,\omega^k_{i}\,p_{}(z_{i}|a^k_{1:i},y_{1:i})\eqsp.
\]
In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by running a Kalman filter as follows. 
\paragraph{Initialization}
At time $i=1$, $(a^k_1)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,J\}$ with probabilities proportional to $(\pi_j)_{1\le j\le J}$. Then, $\mu_1^k$ and $P_1^k$ are computed using a Kalman filter:
\begin{align*}
K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
\end{align*}
Particle $a_1^k$ is associated with the importance weight $\omega^k_1 \propto p(y_1|a_1^k)$. This weight is updated by noting that the conditional distribution of $Y_1$ given $a_1^k$ is a Gaussian distribution with mean $c_{a^k_1}+B_{a^k_1}\mu_1$ and variance $B_{a^k_1}\Sigma_1B'_{a^k_1} + \barG_{a^k_1}$. 
\paragraph{Iterations}
For all $2\le i \le n$, particle indices $(I^k_i)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,N\}$ with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.  Labels trajectories at time $i-1$ are defined by:
\[
a_{1:i-1}^k = (a^{\mathsf{A}^k_1}_1,\ldots,a^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp,
\]
where
\[
\mathsf{A}^k_{i-1} = I^k_i \quad\mbox{and for } 1\le j<i\quad \mathsf{A}^k_{j} = I^{\mathsf{A}^k_{j+1}}_{j+1}\eqsp.
\]
Similarly, means and variances trajectories are defined by: 
\[
\mu_{1:i-1}^k = (\mu^{\mathsf{A}^k_1}_1,\ldots,\mu^{\mathsf{A}^k_{i-1}}_{i-1})\quad\mbox{and}\quad P_{1:i-1}^k = (P^{\mathsf{A}^k_1}_1,\ldots,P^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp.
\]
The new indicator variables $(a_i^k)_{1\le k \le N}$ are sampled independently in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$, where for all $j \in \{1,\ldots, J\}$:
\[
\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j)\eqsp.
\]
The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ which means that $\gamma_i^{j,k}$ may be written:
\[
\gamma_i^{j,k} =  Q(a_{i-1}^k,j)|B_jP^k_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^k_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^k_{i|i-1}}^2\right\} \eqsp,
\]
where 
\begin{align*}
\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp.
\end{align*}
$\mu_i^k$ and $P_i^k$ are updated according to:
\begin{align*}
K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
\end{align*}
The importance weights are given by $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.

\subsection{Path-space smoother of \cite{kitagawa:1996}}
As noted in \cite{kitagawa:1996}, a smoother may be obtained as a byproduct of the forward filter by keeping track of the genealogy of the weighted samples obtained at each time step. In this case the whole trajectories $(a^k_{1:n})_{1\le k\le n}$ are used to approximate the joint smoothing distributions :
\[
p^{\mathsf{Kit}}_N(a_{1:n}|Y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^k_{1:n}}(a_{1:n})\eqsp.
\]
In this case, the marginal smoothing distribution for a given time $i$ is obtained using the genealogical indices $(\mathsf{A}^k_i)_{1\le k \le N}$:
\[
p^{\mathsf{Kit}}_N(a_{i}|Y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^{\mathsf{A}^k_i}_{i}}(a_{i})\eqsp.
\]
This is a  poor approximation of the marginal smoothing distributions for small values of $i$ as the depletion of the particles increases with the successive resampling steps. See Section~\ref{sec:numerical:experiments} for a numerical illustration.

>>>>

\begin{abstract}
This paper focuses on Sequential Monte Carlo based approximations of smoothing distributions in conditionally linear and Gaussian state spaces. Only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps for these models. The algorithm proposed by  \cite{briers:doucet:maskell:2010} extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps. Other Rao Blackwellised Sequential Monte Carlo smoothers introduced in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} are based on the forward backward decomposition of the marginal smoothing distribution% In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps 
 and use explicit marginalization of the linear states in both forward and backward passes. In this paper, we propose an overview of these algorithms and introduce  compared to an extension of the Two-filter based algorithm of \cite{briers:doucet:maskell:2010} with additional explicit Rao Blackwellised steps to improve its performance. The procedures are applied to commodity markets which are described using a two factor model based on the spot price and a convenience yield. Parameters are estimated using  an Expectation Maximization and the performance of the proposed procedure is assessed with Chicago Mercantile Exchange crude oil data.

@@

An important feature of these models is that, conditionally on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian. To exploit this specific structure, it has been suggested in \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to combine Sequential Monte Carlo (SMC) to sample indicator variables with the Kalman filter leading to Rao Blackwellised Monte Carlo filters. %These SMC methods based on sequential importance sampling and resampling techniques introduced in \cite{handschin:mayne:1969} and \cite{rubin:1987} allow to approximate any sequence of conditional distributions of the hidden states given the observations, see \cite{kitagawa:1987,gordon:salmond:smith:1993}. 
Improvements of these early filtering techniques  have been introduced in \cite{doucet:gordon:krishnamurthy:2001,schon:gustafsson:nordlund:2005}. %These algorithms sample only trajectories of indicator variables associated with importance weights, the corresponding state variables $(Z_1,\ldots,Z_k)$ are marginalized and estimated with a Kalman filter, which is the optimal filter in this case. 
%However, many maximum likelihood based estimation procedures such as  the Expectation Maximization  (EM)  algorithm introduced in \cite{dempster:laird:rubin:1977} require to approximate some marginal smoothing distributions of the states and indicator variables. 
%Several algorithms have been developed to solve the smoothing problem using SMC methods: for instance the Forward Filtering Backward Smoothing algorithm (FFBS), the Forward Filtering Backward Simulation algorithm (FFBSi) and the PArticle-based, Rapid Incremental Smoother (PaRIS)  developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000},  \cite{godsill:doucet:west:2004} and \cite{olsson:westerborn:2015}. 
%,lindsten:schon:2011
%In the forward pass of these algorithms, a Rao Blackwellised particle filter such as \cite{chen:liu:2000} is used to approximate the filtering distributions and the corresponding particles, weights and associated trajectories of means and variances of the Kalman filters are stored.
The use of Rao Blackwellisation to solve the smoothing problem has been proved to be challenging and has received satisfactory solutions only recently. The first forward-backward smoother proposed in the literature \cite{fong:godsill:doucet:west:2002} was not fully Rao Blackwellised as it required to sample the hidden linear states in the backward pass. An alternative, based on a structural approximation of the model suggested in \cite{kim:1994}, was introduced by \cite{barber:2006} to handle the backward pass. This approximation introduces a systemic bias which might be difficult to control.  The first fully Rao Blackwellised SMC smoother which should lead to consistent approximations when the number of particles grows to infinity was proposed by  \cite{briers:doucet:maskell:2010} and extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.  This two-filter approach combines a forward and a backward filters which are approximated numerically using SMC for the indicator variables and Kalman filtering techniques for the hidden linear states. More recently, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} introduced a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions. 
%The backward pass samples trajectories of indicator variables backward in time using the particles and weights of the forward pass.

In this paper, we propose to improve the performance of the algorithms introduced in \cite{briers:doucet:maskell:2010} and in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} by using additional Rao Blackwellisation steps, extending the ideas of \cite{fa}.  In \cite{briers:doucet:maskell:2010}, for all $1\le i\le n$, the sampled forward and backward sequences are merged to approximate the posterior distribution of $(a_i,z_i)$. We propose to approximate the marginal smoothing distribution of $(a_i,z_i)$ by merging the sampled forward and backward trajectories at times $i-1$ and $i+1$ and integrating out all possible paths between time $i-1$ and time $i$ and between time $i$ and time $i+1$ instead of sampling random variables. Similarly, in the backward pass of \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013}, an indicator variable $\tilde{a}_i$ is sampled at time $1\le i\le n-1$ using the particles produced by the forward filter at time $i$. In this case, particle rejuvenation may be introduced by using the forward weighted samples at time $i-1$ and extending these trajectories at time $i$ with a Kalman filter for all possible values of the indicator variable. Then, $\tilde{a}_i$ is sampled in $\{1,\ldots,J\}$ using an appropriately adapted weight.

The paper is organized as follows. An extensive survey of the proposed fully Rao Blackwellised Monte Carlo smoothers  as long as the additional explicit marginalization steps are presented in Section~\ref{sec:RaoBlackwell}. The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data. In Section~\ref{}, an application of regime switchings for commodity markets is presented using CME crude oil (WTI) data. A detailed derivation of the algorithms is provided in the Appendix.

%Then, a new model is proposed to explain regime switchings for commodity markets. The parameters of the proposed model are estimated using a Monte Carlo EM algorithm where the required conditional expectations are replaced by their estimates based on the proposed SMC method.  
%The performance of the extended Rao-Blackwellised two-filter  algorithm is assessed using simulated data and CME crude oil (WTI) data.

\section{Rao Blackwellised smoothing algorithms}
\label{sec:RaoBlackwell}
This section details the Sequential Monte Carlo algorithms which can be used to approximate the conditional distribution of the states $(a_1,\ldots,a_n)$ or the marginal distributions of $(a_i,z_{i})$ given the observations $(Y_1,\ldots,Y_n)$. For all $m\,\times\,m$ matrix  let $|A|$ be the determinant of $A$ if $A$ is positive-definite matrix, for all $z\in\mathbb{R}^{m}$ define
\[
\normMat{A}{z}^2 \eqdef z'A^{-1}z\eqsp.
\]
Let $m_{}(a_{i},z_{i-1};z_i)$ be the probability density of the conditional distribution of $Z_i$ given $(a_{i},Z_{i-1})$ and $g_{}(a_{i},z_{i};y_i)$ be the probability density of the conditional distribution of $Y_i$ given $(a_{i},Z_{i})$:
\begin{align}
\label{eq:definition-m}
m_{}(a_{i},z_{i-1};z_i) & \eqdef \left|2\pi\barH_{a_i}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\barH_{a_i}}{z_i -d_{a_{i}} -T_{a_{i}}z_{i-1}}^2\right\}\eqsp,\\
\label{eq:definition-g}
g_{}(a_{i},z_{i};y_i) & \eqdef \left|2\pi\barG_{a_i}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\barG_{a_i}}{y_i - c_{a_i} - B_{a_i}z_{i}}^2\right\}\eqsp,
\end{align}
where
\[
\barG_j \eqdef G_jG'_j\;,\; \barH_{j} \eqdef H_jH'_j\eqsp.
\]
%Maximum likelihood estimation of the parameters using the Expectation Maximization algorithm or gradient ascent algorithms requires to compute the posterior distributions of the states $a_{i-1:i},z_{i-1:i}$  given the observations $Y_{1:n}$, written $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$, which cannot be computed in closed-form (the dependency of $\phi^n_{i}$ on $Y_{1:n}$ is dropped for better clarity). SMC methods provide a class of importance sampling and resampling procedures to approximate $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$. 
All the algorithms considered in this paper are based on forward-backward or two-filter decompositions of the smoothing distributions and share the same forward filter detailed in Section~\ref{sec:forward}.

\subsection{Forward filter}
\label{sec:forward}
The SMC approximation $p^N_{}(a_{1:i},z_{i}|y_{1:i})$ of $p_{}(a_{1:i},z_{i}|y_{1:i})$ may be obtained using the standard Rao Blackwellised  algorithm given in \cite{chen:liu:2000}. The procedure produces a sequence of trajectories $(a^k_{1:i})_{1\le k \le N}$ associated with importance weights $(\omega^k_{i})_{1\le k \le N}$ used to define the following approximation of $p_{}(a_{1:i},z_{i}|y_{1:i})$:
\[
p^N_{}(a_{1:i},z_{i}|y_{1:i}) = \sum_{k=1}^N\delta_{a^k_{1:i}}(a_{1:i})\,\omega^k_{i}\,p_{}(z_{i}|a^k_{1:i},y_{1:i})\eqsp.
\]
In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by using the Kalman filter's update. 
\paragraph{Initialization}
At time $i=1$, $(a^k_1)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,J\}$ with probabilities proportional to $(\pi_j)_{1\le j\le J}$. Then, $\mu_1^k$ and $P_1^k$ are computed using a Kalman filter:
\begin{align*}
K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
\end{align*}
Particle $a_1^k$ is associated with the importance weight $\omega^k_1 \propto p(y_1|a_1^k)$. This weight is updated by noting that the conditional distribution of $Y_1$ given $a_1^k$ is a Gaussian distribution with mean $c_{a^k_1}+B_{a^k_1}\mu_1$ and variance $B_{a^k_1}\Sigma_1B'_{a^k_1} + \barG_{a^k_1}$. 
\paragraph{Iterations}
%For all $2\le i \le n$, particle indices $(I^k_i)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,N\}$ with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.  The ancestral path of indicator variables at time $i-1$ is defined by:
%\begin{equation}
%\label{eq:ancestral:traj}
%a_{1:i-1}^k = (a^{\mathsf{A}^k_1}_1,\ldots,a^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp,
%\end{equation}
%where the ancestral variables are defined as follows:
%\[
%\mathsf{A}^k_{i} = k \quad\mbox{and for } 1\le j<i\quad \mathsf{A}^k_{j} = I^{\mathsf{A}^k_{j+1}}_{j+1}\eqsp.
%\]
%Similarly, means and variances trajectories are defined by: 
%\[
%\mu_{1:i-1}^k = (\mu^{\mathsf{A}^k_1}_1,\ldots,\mu^{\mathsf{A}^k_{i-1}}_{i-1})\quad\mbox{and}\quad P_{1:i-1}^k = (P^{\mathsf{A}^k_1}_1,\ldots,P^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp.
%\]
%The new indicator variables $(a_i^k)_{1\le k \le N}$ are sampled independently in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$, where for all $j \in \{1,\ldots, J\}$:
For all sampled trajectories $\{a_{1:i-1}^k\}_{1\le k \le N}$ and all $1\le j \le J$, define:
\[
\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j)\eqsp.
\]
The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ so that:% which means that $\gamma_i^{j,k}$ may be written:
\[
\gamma_i^{j,k} \propto  Q(a_{i-1}^k,j)|B_jP^{j,k}_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^{j,k}_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^{j,k}_{i|i-1}}^2\right\} \eqsp,
\]
where 
\begin{align*}
\mu^{j,k}_{i|i-1} &= d_{j} + T_{j}\mu^k_{i-1}\eqsp,\\
P^{j,k}_{i|i-1} &= T_{j}P^k_{i-1}T'_{j} + \barH_{j}\eqsp.
\end{align*}

%$\mu_i^k$ and $P_i^k$ are updated according to:
%\begin{align*}
%K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
%\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
%P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
%\end{align*}
%The importance weights are given by $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.

%\subsection{Path-space smoother of \cite{kitagawa:1996}}
%As noted in \cite{kitagawa:1996}, a smoother may be obtained as a byproduct of the forward filter by keeping track of the genealogy of the weighted samples obtained at each time step. In this case the whole trajectories $(a^k_{1:n})_{1\le k\le n}$ are used to approximate the joint smoothing distributions :
%\[
%p^{\mathsf{Kit}}_N(a_{1:n}|y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^k_{1:n}}(a_{1:n})\eqsp.
%\]
%In this case, the marginal smoothing distribution for a given time $i$ is obtained using the genealogical indices $(\mathsf{A}^k_i)_{1\le k \le N}$:
%\[
%p^{\mathsf{Kit}}_N(a_{i}|y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^{\mathsf{A}^k_i}_{i}}(a_{i})\eqsp.
%\]
%This is a  poor approximation of the marginal smoothing distributions for small values of $i$ as the depletion of the particles increases with the successive resampling steps. See Section~\ref{sec:numerical:experiments} for a numerical illustration.

\subsection{Algorithm based on the structural approximation of \cite{kim:1994}, \cite{barber:2006}}
The method to approximate the smoothing distributions of the indicator variables given all the observations proposed by \cite{barber:2006} relies on an tractable approximation of $p(a_i|a_{i+1:n},y_{1:n})$. Assume first that the particles and importance weights $(a_i^k,\omega_i^k)_{1\le k \le N}$, $1\le i\le n$, approximating the filtering distributions are available. Then, note that, for all $1\le i\le n$,
\[
p(a_{i:n}|y_{1:n}) = p(a_{i}|a_{i+1:n},y_{1:n})p(a_{i+1:n}|y_{1:n})\eqsp.
\]
Then, the approximation of \cite{kim:1994,barber:2006} replaces $p(a_i|a_{i+1:n},y_{1:n})$ by
\[
p(a_i|a_{i+1},y_{1:i}) \propto Q(a_i,a_{i+1})p(a_i|y_{1:i})\eqsp.
\]
Plugging the particle approximation of $p(a_i|y_{1:i})$ in this expression provides the following procedure to sample trajectories approximatively distributed according to $p(a_{1:n}|y_{1:n})$:
\begin{enumerate}[-]
\item $\tilde{a}_{n}$ is first sampled in $\{a_n^1,\ldots,a_n^N\}$ with probabilities proportional to $(\omega_n^k)_{1\le k \le N}$. 
\item Then, for all $1\le i \le n-1$, $\tilde{a}_{i}$ is sampled in $\{a_i^1,\ldots,a_i^N\}$ with probabilities proportional to $(\omega_i^kQ(a_i^k,\tilde{a}_{i+1}))_{1\le k \le N}$. 
\end{enumerate}
If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{kim:1994} of the joint smoothing distribution of the indicator variables is:
\[
p^{\mathsf{Kim}}_N(a_{1:n}|Y_{1:n}) = \frac{1}{\tilde{N}}\sum_{k=1}^{\tilde N} \delta_{\tilde{a}^{k}_{1:n}}(a_{1:n})\eqsp.
\]
